|

### 2020-07-03 paper.Rmd
US spelling.

The notion of a "curse of dimensionality" was originally introduced in @BellmanRichard1961 to describe the exponentially growing difficulty of optimisation over an increasing number of dimensions. More generally, we can think of it as a problem of sampling large parameter spaces. As the number of dimensions increases, the volume of the parameter space grows exponentially, and the density of points decreases. As a result observations will be far from the center of the distribution, and in the extreme case of high-dimension low-sample size date observations are on the vertices of a simplex [@doi:10.1111/j.1467-9868.2005.00510.x]. This causes problems in different domains. An important example in statistics is cluster analysis. Since the distance between any two points in a distribution becomes approximately constant, new methods are needed to cluster points in a high-dimensional space (see e.g. @Steinbach2004).

A more detailed discussion of the curse of dimensionality was presented in @Donoho00, which also highlighted the opposite effect, a blessing of dimensionality. Indeed we can sometimes exploit large parameter spaces to our advantage. One example is the use of regularisation methods for variable selection. These methods penalise model complexity in the optimisation, as for example in ridge regression and Lasso. The penalty results in shrinking (some of) the parameter estimates towards zero. In ridge regression we obtain a global shrinkage, while Lasso can be thought of as variable selection and will shrink some of the estimates to zero. XXX mention concentration of measures?

A different view on the curse of dimensionality emerges for dimension reduction and visualisation methods. When working with low-dimensional linear projections from a high-dimensional parameter space, most projections will show approximately Gaussian distributions of the data, and observations start to pile near the center as dimensionality increases [@diaconis1984]. For high-dimension low-sample size data this results in "data piling", i.e. we can find directions along which all observations are in a single point [@10.2307/27639976,@10.1093/biomet/asp084]. These issues also persist with non-linear dimension reduction techniques. This is often referred to as the "crowding problem", which methods like t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne] aim to alleviate.

To understand these seemingly contradictory implications, low density of points near the center in the high-dimensional space, and the piling in low-dimensional projections, we need to consider the projected volume across the low-dimensional projection. We can think of the data being distributed in a hypersphere, i.e. we think of data points within a specified distance from the center. Compared to the common assumption of a distribution in a box or hypercube this has the advantage of being rotation invariant, and more accurately captures commonly observed distributions. We can picture the linearly projected volume starting from a 3D sphere. In a 2D projection, a large fraction of the volume will be projected onto a small region near the center of the resulting disc, and will be even further concentrated when projecting onto 1D. The same intuition holds as we increase the number of original dimensions that are projected down to one or two dimensions for visualisation, with increasing fractions of the volume being projected onto a small area around the center. The piling of observaitions then follows directly by assuming that they are uniformly distributed across the volume.

@@

As discussed in the introduction, the aim of the radial transformation is to redistribute the projected points, such that equal volume in the original ($p$-dimensional) space is projected onto equal area in a 2-dimensional projection. Note that it is straightforward to generalise this for $d$-dimensional projections by mapping onto equal $d$-dimensional volume instead.

@@

We compare $V^{rel}_{2D} (r, p, R)$ to the relative volume within a radius $r$ in the full hypersphere,
\begin{equation}
V^{rel}_{pD} (r, p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p,
\end{equation}
to illustrate the opposing effects on their distribution. This comparison is shown in Fig. \ref{fig:cdf}, for $p=3, 10, 100$. The left graph shows $V^{rel}_{pD} (r, p, R)$ and the right graph $V^{rel}_{2D} (r, p, R)$. We clearely see the opposing behaviour of the two functions as $p$ increases. While $V^{rel}_{pD} (r, p, R)$ concentrates more and more towards large values of $r/R$, $V^{rel}_{2D} (r, p, R)$ has the opposite behaviour. This concentration near the center of the disk is the result of integrating over all orthogonal directions when projecting onto the plane, and should be reversed by our radial transformation of the projected points.

@@

The relation between $r'$ and $r$ depends on the number of dimensions $p$, and is illustrated for selected values in Figure \ref{fig:radii}. We see that the transformation is approximately linear near the center. As $p$ increases it becomes non-linear faster, and e.g. for $p=10$ the points with radius $r>0.5$ will alrady be highly distorted and pushed out towards the last eighth in $r'$. Figure \ref{fig:circles} demonstrates this for different values of $p$ by showing equidistant circles for which the radius has been transformed according to Eq. \ref{eq:resc}.

@@

For example, the dimension of the input might not reflect the intrinsic dimensionality of the dataset. This could be the case when dimension reduction was used prior to visualisation, e.g. displaying only the first few principal components. In this case the preferred effective dimensionality $p_{eff}$ is likely between the original numbr of dimensions and the selected number of principal components, since some directions would be considered pure noise, while others might still carry relevant information.

@@

In practice we allow tuning $p_{eff} = s p$ by selecting the scaling parameter $s$. By default, $s=1$ and $p_{eff}=p$. When $s<1$ the rescaling will be softer, and $s>1$ results in more aggressive rescaling than suggested by $p$ alone. Note that when $p_{eff} < 2$ we actually invert the behaviour and shift the focus away from the center, in general this is not wanted.

@@

We can use the new display to look at different distribution of images from Google quickdraw (XXX REFERENCE). These are $28\times28=784$ pixel greyscale data, large dataset publicly available, here use small subset of three types of sketches (banana, cactus, crab) and observations (1000 sketches from each class) and see if we can separate the classes in the high-dimensional parameter space. We first reduce dimensionality using PCA and look at the first five PCs.

>>>>

The notion of a "curse of dimensionality" was originally introduced in @BellmanRichard1961 to describe the exponentially growing difficulty of optimization over an increasing number of dimensions. More generally, we can think of it as a problem of sampling large parameter spaces. As the number of dimensions increases, the volume of the parameter space grows exponentially, and the density of points decreases. As a result observations will be far from the center of the distribution, and in the extreme case of high-dimension low-sample size date observations are on the vertices of a simplex [@doi:10.1111/j.1467-9868.2005.00510.x]. This causes problems in different domains. An important example in statistics is cluster analysis. Since the distance between any two points in a distribution becomes approximately constant, new methods are needed to cluster points in a high-dimensional space (see e.g. @Steinbach2004).

A more detailed discussion of the curse of dimensionality was presented in @Donoho00, which also highlighted the opposite effect, a blessing of dimensionality. Indeed we can sometimes exploit large parameter spaces to our advantage. One example is the use of regularization methods for variable selection. These methods penalize model complexity in the optimization, as for example in ridge regression and Lasso. The penalty results in shrinking (some of) the parameter estimates towards zero. In ridge regression we obtain a global shrinkage, while Lasso can be thought of as variable selection and will shrink some of the estimates to zero. XXX mention concentration of measures?

A different view on the curse of dimensionality emerges for dimension reduction and visualization methods. When working with low-dimensional linear projections from a high-dimensional parameter space, most projections will show approximately Gaussian distributions of the data, and observations start to pile near the center as dimensionality increases [@diaconis1984]. For high-dimension low-sample size data this results in "data piling", i.e. we can find directions along which all observations are in a single point [@10.2307/27639976,@10.1093/biomet/asp084]. These issues also persist with non-linear dimension reduction techniques. This is often referred to as the "crowding problem", which methods like t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne] aim to alleviate.

To understand these seemingly contradictory implications, low density of points near the center in the high-dimensional space, and the piling in low-dimensional projections, we need to consider the projected volume across the low-dimensional projection. We can think of the data being distributed in a hypersphere, i.e. we think of data points within a specified distance from the center. Compared to the common assumption of a distribution in a box or hypercube this has the advantage of being rotation invariant, and more accurately captures commonly observed distributions. We can picture the linearly projected volume starting from a 3D sphere. In a 2D projection, a large fraction of the volume will be projected onto a small region near the center of the resulting disc, and will be even further concentrated when projecting onto 1D. The same intuition holds as we increase the number of original dimensions that are projected down to one or two dimensions for visualization, with increasing fractions of the volume being projected onto a small area around the center. The piling of observations then follows directly by assuming that they are uniformly distributed across the volume.

@@

As discussed in the introduction, the aim of the radial transformation is to redistribute the projected points, such that equal volume in the original ($p$-dimensional) space is projected onto equal area in a 2-dimensional projection. Note that it is straightforward to generalize this for $d$-dimensional projections by mapping onto equal $d$-dimensional volume instead.

@@

We compare $V^{rel}_{2D} (r, p, R)$ to the relative volume within a radius $r$ in the full hypersphere,
\begin{equation}
V^{rel}_{pD} (r, p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p,
\end{equation}
to illustrate the opposing effects on their distribution. This comparison is shown in Fig. \ref{fig:cdf}, for $p=3, 10, 100$. The left graph shows $V^{rel}_{pD} (r, p, R)$ and the right graph $V^{rel}_{2D} (r, p, R)$. We clearly see the opposing behavior of the two functions as $p$ increases. While $V^{rel}_{pD} (r, p, R)$ concentrates more and more towards large values of $r/R$, $V^{rel}_{2D} (r, p, R)$ has the opposite behavior. This concentration near the center of the disk is the result of integrating over all orthogonal directions when projecting onto the plane, and should be reversed by our radial transformation of the projected points.

@@

The relation between $r'$ and $r$ depends on the number of dimensions $p$, and is illustrated for selected values in Figure \ref{fig:radii}. We see that the transformation is approximately linear near the center. As $p$ increases it becomes non-linear faster, and e.g. for $p=10$ the points with radius $r>0.5$ will already be highly distorted and pushed out towards the last eighth in $r'$. Figure \ref{fig:circles} demonstrates this for different values of $p$ by showing equidistant circles for which the radius has been transformed according to Eq. \ref{eq:resc}.

@@

For example, the dimension of the input might not reflect the intrinsic dimensionality of the dataset. This could be the case when dimension reduction was used prior to visualization, e.g. displaying only the first few principal components. In this case the preferred effective dimensionality $p_{eff}$ is likely between the original number of dimensions and the selected number of principal components, since some directions would be considered pure noise, while others might still carry relevant information.

@@

In practice we allow tuning $p_{eff} = s p$ by selecting the scaling parameter $s$. By default, $s=1$ and $p_{eff}=p$. When $s<1$ the rescaling will be softer, and $s>1$ results in more aggressive rescaling than suggested by $p$ alone. Note that when $p_{eff} < 2$ we actually invert the behavior and shift the focus away from the center, in general this is not wanted.

@@

We can use the new display to look at different distribution of images from Google quickdraw (XXX REFERENCE). These are $28\times28=784$ pixel grey scale data, large dataset publicly available, here use small subset of three types of sketches (banana, cactus, crab) and observations (1000 sketches from each class) and see if we can separate the classes in the high-dimensional parameter space. We first reduce dimensionality using PCA and look at the first five PCs.
